Jane, John … Leslie? A Historical Method for Algorithmic Gender Detection

### 2014-09-15 article.Rmd
Citations

But researchers studying a longer timespan need to take into account the changing nature of naming practices. If that same education researcher wanted to compare the demographics of contemporary urban schools with those from the early twentieth century, contemporary tools like Genderize.io would misidentify Leslie Fay's gender in historical datasets. The "Leslie" problem is not just for researchers who think of themselves as historians. Anyone studying a period longer than a few years, or anyone studying a group whose demographics do not match those of a contemporary tool like Genderize.io will also encounter this problem. As of 2012, the average American has a lifespan of nearly seventy-nine years–--more than enough time for naming practices to change quite dramatically between their birth and death.^[World Bank, *World Bank Open Data*, "Life expectancy at birth, total (years)," <http://data.worldbank.org/indicator/SP.DYN.LE00.IN> (accessed September 12, 2014).] Predicting gender from first names therefore requires a fundamentally historical method.

Our solution to the "Leslie" problem is to create a software package that combines a predictive algorithm with a several historical datasets suitable to various times and regions. The algorithm calculates the proportion of male or female uses of a name in a given birth year or range of years. It thus can provide not only a prediction of the gender of a name, but also an estimate of how likely that prediction is. For example 52 percent of the babies named Leslie born in the United States in 1950 were girls and 48 percent were boys. This allows researchers to conduct more nuanced analysis and to determine for themselves what proportion is acceptable for making a prediction of gender.

We recognize that gender is a fluid concept, both today and in the past, and that this method glosses over the complexities of historical definitions of gender and of the relationship between biological sex and gender. In some sense this is unavoidable since the large historical datasets which make this kind of analysis possible were created by state agencies. For the category of gender (as for race) states have offered a limited number of boxes to check, and those boxes extert a tremendous power to define state subjects.^[One of the best-known theorists on gender is Judith Butler, *Gender Trouble* and Butler, *Bodies that Matter*; For the problems with the word "identity" in the humanities and social sciences, see Rogers Brubaker and Frederick Cooper, "Beyond 'Identity'"; Several historians have written about the American state and its drive to impose identities onto subjects: Margot Canady, The Straight State;  Peggy Pascoe, *What Comes Naturally*; Shane Landrum on birth certificates.] Nevertheless, this method is neutral as to what the categories male or female actually mean in any given historical context. Researchers are free---indeed, obligated---to make that determination for themselves, and this package provides access to the underlying dataset so that researchers can interrogate every assumption that comes with them. This method does not argue against the complexities of historical understandings of gender, but as with any project it does require researchers who use the method to think through how the data was gathered, what it implies, and how it reflects social and cultural practices.

@@

Why do scholars need a historical method for inferring gender? Let's start with a comparison. One existing method for predicting gender is available in the [Natural Language Toolkit](http://www.nltk.org/) for Python.^[Steven Bird, Edward Loper, and Ewan Klein, *Natural Language Processing with Python* (O’Reilly, 2009), ch. 2. Chapter 6 of the NLTK book also describes a method of predicting gender from the last letter of the name, but this example is given for the sake of explaining supervised classification and not as an end in itself, so we do not have reason to address this a historical method.] The NLTK is an influential software package for scholarship because it provides an extraordinary range of tools for analyzing natural language.^[The NLTK book (the most common way to cite the software) has been cited 812 times according to [Google Scholar](http://scholar.google.com/scholar?cites=3735380965977150091&as_sdt=5,47&sciodt=0,47&hl=en). Because in the humanities (unlike in other fields) the practice of citing even software which determines the outcome of an analysis is regrettably rare, even nonexistent, such metrics are difficult to determine. Yet searching of journals such as *Digital Humanities Quarterly* and the *Journal of Digital Humanities* reveals that natural language processing, for which the NLTK is the foremost implementation, is one of the key methodological categories of the digital humanities.] Included in the NLTK are two lists of male and female names created by the computer scientist Mark Kantrowitz with assistance from Bill Ross.^[The [original lists of names](http://www.cs.cmu.edu/afs/cs/project/ai-repository/ai/areas/nlp/corpora/names/0.html) are available at the website of the computer science department at Carnegie Mellon University, and the NLTK implementation is described in its [documentation](http://www.nltk.org/howto/corpus.html).] These lists include `r kantrowitz_unique` unique names. Using the Kantrowitz names corpus in the NLTK, one could look up a name like Jane or John and find out that it is male or female. The Kantrowitz corpus provides the list of names.^[The examples below is drawn from our Gender package which implements the Kantrowitz dataset for comparsion's sake, though its use is not recommended for actual research. The Gender package combines the Kantrowitz corpus's two lists into a single list, identifying names that are ambiguous because they appear on both lists as `"either"`.]

@@

We mention the Kantrowitz name corpus as implemented in NLTK because the Natural Language Toolkit is rightly regarded as influential for scholarship. Its flaws are also typical of the problems with most other implementations of gender prediction algorithms. The Genderize.io API is, for example, a more sophisticated implementation of gender prediction than the NLTK algorithm. Besides predicting male or female for gender, it also reports the proportion of male or female names, along with a count of the number of uses of the name that it is basing its prediction on. Genderize will also permit the user to customize a prediction for different countries, which is an extremely important feature. Genderize.io reports that its "database contains 142848 distinct names across 77 countries and 85 languages." Genderize is unsuitable for historical work, however, because it is based only on contemporary data. According to the documentation for its API, "It utilizes big datasets of information, from user profiles across major social networks."^[Quotations are taken from the [Genderize.io website](http://genderize.io), with correspondence from the developer behind the site for confirmation.] It would be anachronistic to apply these datasets to the past, and Genderize.io provides no functionality to filter results chronologically as it does geographically. In addition, Genderize.io does not make clear exactly what comprises the dataset and how it was gathered, which keeps scholars from interrogating the value of the source.

R and Python are the most commonly used languages for data analysis among digital humanists.^[Surveys and analysis of useage point out the growth of R and the continuing popularity of Python among "data science" generally. See, for example, Robert A. Muenchen, "The Popularity of Data Analysis Software," *r4stats.com*, <http://r4stats.com/articles/popularity/> (accessed September 12, 2014); Gregory Piatetsky, "Top Languages for analytics, data mining, data science," *KDN Nuggets*, <http://www.kdnuggets.com/2013/08/languages-for-analytics-data-mining-data-science.html> (accessed September 12, 2014).] Python's existing packages for gender prediction all implement a method similar to the NLTK or Genderize.io.^[Python library names] To our knowledge, [CRAN](http://cran.rstudio.com/) (the central repository for R packages) does not include any packages dedicated to gender prediction. In two of the most popular data analysis languages in the digital humanities there is currently no satisfactory existing method of gender prediction for historical and humanistic research.

@@

@@

@@

>>>>

But researchers studying a longer timespan need to take into account the changing nature of naming practices. If that same education researcher wanted to compare the demographics of contemporary urban schools with those from the early twentieth century, contemporary tools like Genderize.io would misidentify Leslie Fay's gender in historical datasets. The "Leslie" problem is not just for researchers who think of themselves as historians. Anyone studying a period longer than a few years, or anyone studying a group whose demographics do not match those of a contemporary tool like Genderize.io will also encounter this problem. As of 2012, the average American has a lifespan of nearly seventy-nine years–--more than enough time for naming practices to change quite dramatically between their birth and death [World Bank 2014]. Predicting gender from first names therefore requires a fundamentally historical method.

Our solution to the "Leslie" problem is to create a software package that combines a predictive algorithm with several historical datasets suitable to various times and regions. The algorithm calculates the proportion of male or female uses of a name in a given birth year or range of years. It thus can provide not only a prediction of the gender of a name, but also a measure of that prediction's likelihood. For example, our program predicts that a person named Leslie born in 1950 is female, but only just barely: 52 percent of the babies named Leslie born in the United States in 1950 were girls and 48 percent were boys. Including these proportion figures allows researchers to conduct more nuanced analysis and to determine for themselves what proportion is acceptable predicting the gender of names.

We recognize that gender is a fluid concept, both today and in the past, and that this method glosses over the complexities of historical definitions of gender and of the relationship between biological sex and gender \[Butler 1990, Butler 1993]\. In some sense this is unavoidable since the large historical datasets which make this kind of analysis possible were created by state agencies. For the category of gender (as for race) states have offered a limited number of boxes to check, and those boxes extert a tremendous power to define state subjects \[Pascoe 2009, Canaday 2011, Landrum 2014]\.^[The word "identity" is a loaded one in the humanities and social sciences \[Brubaker and Cooper 2000]\. ] Nevertheless, this method is neutral as to what the categories male or female actually mean in any given historical context. Researchers are free---indeed, obligated---to make that determination for themselves, and this package provides access to the underlying dataset so that researchers can interrogate every assumption that comes with them. This method does not argue against the complexities of historical understandings of gender, but as with any project it does require researchers who use the method to think through how the data was gathered, what it implies, and how it reflects social and cultural practices.

@@

Why do scholars need a historical method for inferring gender? Let's start with a comparison. One existing method for predicting gender is available in the [Natural Language Toolkit](http://www.nltk.org/) for Python \[Bird et al. 2009, Chapter 2]\.^[Chapter 6 of \[Bird et al. 2009] also describes a method of predicting gender from the last letter of the name, but this example is given for the sake of explaining supervised classification and not as an end in itself, so we do not have reason to address this a historical method.] The NLTK is an influential software package for scholarship because it provides an extraordinary range of tools for analyzing natural language.^[\[Bird et al. 2009]\ has been cited 812 times according to [Google Scholar](http://scholar.google.com/scholar?cites=3735380965977150091&as_sdt=5,47&sciodt=0,47&hl=en). Because in the humanities (unlike in other fields) the practice of citing software (even software which determines the outcome of an analysis) is regrettably rare, even nonexistent, such metrics are difficult to determine. Yet searching journals such as *Digital Humanities Quarterly* and the *Journal of Digital Humanities* reveals that natural language processing, for which the NLTK is the foremost implementation, is one of the key methodological categories of the digital humanities.] Included in the NLTK are two lists of male and female names created by the computer scientist Mark Kantrowitz with assistance from Bill Ross.^[The [original lists of names](http://www.cs.cmu.edu/afs/cs/project/ai-repository/ai/areas/nlp/corpora/names/0.html) are available at the website of the computer science department at Carnegie Mellon University, and the NLTK implementation is described in its [documentation](http://www.nltk.org/howto/corpus.html).] These lists include `r kantrowitz_unique` unique names. Using the Kantrowitz names corpus in the NLTK, one could look up a name like Jane or John and find out that it is male or female. The Kantrowitz corpus provides the list of names.^[The examples below is drawn from our Gender package which implements the Kantrowitz dataset for comparsion's sake, though its use is not recommended for actual research. The Gender package combines the Kantrowitz corpus's two lists into a single list, identifying names that are ambiguous because they appear on both lists as `"either"`.]

@@

We mention the Kantrowitz name corpus as implemented in NLTK because the Natural Language Toolkit is rightly regarded as influential for scholarship. Its flaws are also typical of the problems with most other implementations of gender prediction algorithms. The Genderize.io API is, for example, a more sophisticated implementation of gender prediction than the NLTK algorithm. Besides predicting male or female for gender, it also reports the proportion of male or female names, along with a count of the number of uses of the name that it is basing its prediction on. Genderize will also permit the user to customize a prediction for different countries, which is an extremely important feature. Genderize.io reports that its "database contains 142848 distinct names across 77 countries and 85 languages." Genderize is unsuitable for historical work, however, because it is based only on contemporary data. According to the documentation for its API, "It utilizes big datasets of information, from user profiles across major social networks." It would be anachronistic to apply these datasets to the past, and Genderize.io provides no functionality to filter results chronologically as it does geographically. In addition, Genderize.io does not make clear exactly what comprises the dataset and how it was gathered, which keeps scholars from interrogating the value of the source \[Genderize.io 2014-]\.^[Quotations are taken from the [Genderize.io website](http://genderize.io), with correspondence from the developer behind the site for confirmation.]

R and Python are two of the most commonly used languages for data analysis \[Piatesky 2014, Muenchen 2014-]\.^[Surveys and analysis of useage point out the growth of R and the continuing popularity of Python among "data science" generally. See, for example, ] Python's existing packages for gender prediction all implement a method similar to the NLTK or Genderize.io.^[Python library names] To our knowledge, [CRAN](http://cran.rstudio.com/) (the central repository for R packages) does not include any packages dedicated to gender prediction. In two of the most popular data analysis languages in the digital humanities there is currently no satisfactory existing method of gender prediction for historical and humanistic research.

@@

Bird et al. 2009. Bird, Steve, Edward Loper, and Ewan Klein, *Natural Language Processing with Python* (O’Reilly, 2009)

Brubaker and Cooper 2000. Brubaker, Rogers, and Frederick Cooper. “Beyond ‘identity.’” Theory and Society 29.1 (2000): 1–47.

Butler 1989. Butler, Judith. *Gender Trouble: Feminism and the Subversion of Identity*. New York: Routledge, 1990.

Butler 1993. Butler, Judith. *Bodies That Matter: On the Discursive Limits of Sex*. New York: Routledge, 1993.

Canaday 2011. Canaday, Margot. *The Straight State: Sexuality and Citizenship in Twentieth-Century America*. Princeton, N.J.: Princeton University Press, 2011.

Genderize.io 2014-. *Genderize.io*. Accessed 15 September 2014, http://genderize.io/.

@@

Muenchen 2014-. Muenchen, Robert A. "The Popularity of Data Analysis Software," *r4stats.com*. Accessed 12 September 2014, http://r4stats.com/articles/popularity/.

Pascoe 2009. Pascoe, Peggy. *What Comes Naturally: Miscegenation Law and the Making of Race in America*. Oxford University Press, USA, 2009.

Piatetsky 2013. Piatetsky, Gregory. "Top Languages for analytics, data mining, data science," *KDN Nuggets*. 27 August 2013, http://www.kdnuggets.com/2013/08/languages-for-analytics-data-mining-data-science.html.

@@

World Bank 2014-. World Bank. "Life expectancy at birth, total (years)." *World Bank Open Data*. Accessed 12 September 2014, http://data.worldbank.org/indicator/SP.DYN.LE00.IN.
